### 2014-09-05 text/Abstract.tex
Equational theory (Chapter 5) and Abstract

% Abstract

\pdfbookmark[1]{Abstract}{Abstract} % Bookmark name visible in a PDF viewer

@@

\chapter*{Abstract} % Abstract name
The development of programming languages needs to reflect important changes in the industry.
In recent years, this included \eg~the development of parallel programming models (in reaction 
to the multi-core revolution). This thesis is a response to another such revolution -- the 
diversification of devices and systems where programs run.

We develop the foundations for statically typed functional languages that understand 
the \emph{context} or environment in which programs execute. Such context includes different 
platforms (and their versions) in cross-platform applications, resources available in different 
execution environments (\eg \; GPS sensor on a phone and database on the server), but also more 
traditional notions such as variable context (tracking variable usage in static analyses) or 
past values in stream-based data-flow programming.

The thesis presents three \emph{coeffect} calculi that capture different notions of 
context-awareness: \emph{flat} calculus capturing contextual properties of the execution 
environment, \emph{structural} calculus capturing contextual properties related to 
variable usage and \emph{meta-language} calculus which allows reasoning about multiple notions
of context-dependence in a single language and provides pathway to embedding the other two
calculi in existing languages.

Although the focus of this thesis is on the syntactical properties of the presented 
systems, we also discuss their category-theoretical motivation. We introduce the notion of
an \emph{indexed} comonad (the dual of monads) and show how they provide semantics of the
presented three calculi.

>>>>

%!TEX root = ../main.tex
\pdfbookmark[1]{Abstract}{Abstract} % Bookmark name visible in a PDF viewer

@@

\chapter*{Abstract} % Abstract name
The development of programming languages needs to reflect important changes in the way
programs execute. In recent years, this included \eg~the development of parallel programming 
models (in reaction to the multi-core revolution) or improvements in data access technologies. 
This thesis is a response to another such revolution -- the diversification of devices and 
systems where programs run.

The key point made by this thesis is the realization that execution environment or
\emph{context} is fundamental for writing modern applications and that programming 
languages should provide abstractions for programming with context and verifying how 
it is accessed.

We identify a number of program properties that were not connected before, but model some notion 
of context. Our examples include tracking different execution platforms (and their versions) 
in cross-platform development, resources available in different execution environments (\eg~GPS
sensor on a phone and database on the server), but also more traditional notions such as 
variable usage (\eg~in liveness analaysis) or past values in stream-based data-flow programming.

Our first contribution is the discovery of the connection between the above examples and
their novel presentation -- in the form of calculi with type systems and formal semantics --
that highlights this relationship. Our second and third contribution are two \emph{coeffect}
calculi that capture commonalities in the presented examples. In particular, our 
\emph{flat} coeffect calculus models languages with contextual properties of the execution 
environment and our \emph{structural} coeffect calculus models languages where the contextual
properties are attached to the variable usage.

Although the focus of this thesis is on the syntactical properties of the presented 
systems, we also discuss their category-theoretical motivation. We introduce the notion of
an \emph{indexed} comonad (the dual of monads) and show how they provide semantics of the
presented two calculi.

### 2014-09-05 text/Chapter05.tex
Equational theory (Chapter 5) and Abstract

??

@@

\subsection{Properties of reductions}
\label{sec:structural-metatheory}
Similarly to the flat version, the \clstr~calculus is defined abstractly. We cannot define its
operational meaning, because that will differ for every concrete application. For example,
when tracking array accesses, variables are interpreted as arrays and $\pastval{a}{n}$ denotes
access to a specified element.

Just like previously, we can state general properties of the reductions. As the syntax of expressions
is the same for \clstr~as for \clflt, the substitution and reduction $\twoheadrightarrow_\beta$ 
are also the same and can be found in Figure~\ref{fig:flat-substred}.

The structural coeffect calculus \clstr~associates information with individual variables. This means
that when an expression requires certain context, we know from what scope it comes -- the
context must be provided by a scope that defines the associated variable, which is either a
lambda abstraction or global scope. This distinguishes the structural system from the flat system
where context could have been provided by any scope and the lambda rule allowed arbitrary 
splitting of context requirements between the two scopes (or declaration and caller site).

>>>>

Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
different notion of context. As before, the properties of the flat coeffect algebra guarantee that 
certain equational properties hold for all instances of the calculus.

\subsection{From flat to structural coeffects}

The discussion of equational theory for flat calculus in Section~\ref{sec:flat-syntax} considered
different reductions that can be used as a basis for operational semantics for calculi satisfying
different additional properties, as none of the techniques works universally. The structural 
coeffect calculus has more desirable equational properties -- in particular, both $\beta$-reduction
and $\eta$-expansion can be proved using just the properties of strucutral coeffect algebra. Thus
this section focuses on these reductions. 
Using the terminology of Pfenning and Davies \cite{logic-modal-reconstruction},
the structural coeffect calculus satisfies both are \emph{local soundness} and \emph{local completeness}.

% The structural coeffect calculus clstr~associates information with individual variables. This means
% that when an expression requires certain context, we know from what scope it comes -- the
% context must be provided by a scope that defines the associated variable, which is either a
% lambda abstraction or global scope. This distinguishes the structural system from the flat system
% where context could have been provided by any scope and the lambda rule allowed arbitrary 
% splitting of context requirements between the two scopes (or declaration and caller site).

\subsection{Holes substitution lemma}

if

\begin{equation*}
\coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s
\end{equation*}

and

\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_s}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
  \coctx{x_k\!:\!\tau_s}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
\end{equation*}

then

\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{r}}} \,|\, \ldots \,|\, 
  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{r}}}\,] \vdash \subst{\subst{e_r}{x_1}{e_s}\ldots}{x_k}{e_s} : \tau_r
\end{equation*}

\subsection{Structural coeffect systems}
\label{prop:structural}

For structural coeffect systems, recall that coeffects are vectors
with $\czip = \cpar = \cons$ (vector concatenation) and $\czero =
\langle{}\rangle$ (the empty vector), thus coeffect annotations
comprise the \emph{free monoid} \ie, lists over the coeffect scalars
(although we continue using the vector terminology). We first show
substitution:

\begin{lemma}[Substitution lemma] 
\label{thm:structural-subst}
In a structural coeffect calculus with a coeffect scalar structure $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$: 
\[
\begin{array}{l}
\coctx{\Gamma}{\aclrd{S}} \vdash e_s : \sigma  
\;\;  \wedge  \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{\aclrd{R_1} \cons\; \alift{\cclrd{r}} \cons\; \aclrd{R_2} } \vdash e_r : \tau\\
\quad \Rightarrow \; \coctx{\Gamma_1,\Gamma,\Gamma_2}{\aclrd{R_1} \cons\; (\cclrd{r} \,\aseq\, \aclrd{S}) \cons\; \aclrd{R_2}} \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\label{lemm:structural-subst}
\end{lemma}
\begin{proof}
By induction over the $e_2$ derivation using the free monoid structure ($\C, \cons, \langle{}\rangle$)
and coeffect scalar axioms (full proof~\cite{appendix}). %Full proof in the Appendix.
\end{proof}

\noindent
Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
Using the substitution lemma, we can demonstrate $\beta$-equality:
%
\begin{equation*}
\inference{
  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
\equiv
\subst{e_1}{x}{e_2} : \tau}
\end{equation*}
%%
\noindent
As a result, $\beta$-reduction preserves the type and coeffects of a term.
This gives the following subject reduction property:

\begin{theorem}[Subject reduction]
In a structural coeffect calculus, if $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$ and
$e \reduce_\beta e'$ then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
Following from Lemma~\ref{thm:structural-subst} and $\beta$-equality.  
\end{proof}

\noindent
Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
\emph{local soundness} and \emph{local completeness} conditions set by Pfenning and Davies
\cite{logic-modal-reconstruction}. This means that abstraction does not introduce too much, 
and application does not eliminate too much.
%
\begin{equation*}
\inference{
  \inference 
    {\coctx{\Gamma}{\aclrd{R}} \vdash e : \sigma \xrightarrow{s} \tau & 
     \coctx{x:\sigma}{\alift{\cunit}} \vdash x : \sigma}
    {\coctx{\Gamma, x:\sigma}{\aclrd{R} \,\cons\, (\cclrd{s} \,\aseq\, \alift{\cunit})} \vdash e \, x : \tau}}
  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x.e~x  \equiv e 
: \sigma \xrightarrow{\cclrd{s}} \tau}
\end{equation*}
%
The last step uses the fact that 
$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.

This highlights another difference between coeffects and effects, as $\eta$-equality does
not hold for many notions of effect. For example, in a language with output
effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ has different effects to 
its $\eta$-converted form $\lambda x . e x$ because the immediate effects of $e$ are
hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, the (abs) rule
allows immediate contextual requirements of $e$ to ``float outside'' of the enclosing
$\lambda$. Furthermore, the free monoid nature of $\cons$ in structural systems allows 
the exact immediate requirements of $\lambda x . e x$ to match those of $e$.

@@
